---
title: "T02 Descriptive Statistics and Visualization"
subtitle: "MATH	2411 Applied Statistics"
author: "WANG Zhiwei"
institute: "MATH, HKUST"
date: "2024-09-16"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  results = "asis",
  echo = TRUE,
  comment = "#>",
  out.width = "100%"
)
library(xaringanthemer)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
style_mono_accent(
  header_font_google = google_font("Josefin Slab", "600"),
  text_font_google   = google_font("Work Sans", "300", "300i"),
  code_font_google   = google_font("IBM Plex Mono")
)

style_mono_light(
  base_color = "#003366",
  link_color = "#996600",
  text_bold_color = "#996600",
  link_decoration = "underline"
)
```

```{r xaringan-panelset, echo=FALSE}
xaringanExtra::use_panelset()
```

class: inverse, center, middle

# Let's first simulate a dataset!

---

## Student grading system

```{r}
set.seed(20240916)
n_students <- 100; n_groups <- 5
each_group <- n_students / n_groups
id <- as.integer(c(1:n_students))
gender <- as.factor(sample(c("M", "F"), n_students, replace = TRUE))
group <- as.factor(rep(1:n_groups, each_group))
math <- round(runif(min = 55, max = 95, n = n_students))
english <- round(runif(min = 55, max = 95, n = n_students))
physics <- round(runif(min = 55, max = 95, n = n_students))
grading_system <- data.frame(id, group, gender, math, english, physics)
grading_system$average <- rowMeans(grading_system[, c("math", "english", "physics")])
knitr::kable(head(grading_system, n = 3), format = 'html')
```

---

## Some simple descriptive statistics

```{r}
dim(grading_system)
```

--

```{r}
nrow(grading_system)
n_students
```

--

```{r}
ncol(grading_system)
```

---

## Some simple descriptive statistics

```{r}
table(grading_system$gender)
table(grading_system$group)
```

--

```{r}
summary(grading_system$gender)
summary(grading_system$math)
```

---

## Some simple descriptive statistics

```{r}
knitr::kable(summary(grading_system[, c("math", "english", "physics", "average")]), format = 'html')
```

---

## Some simple descriptive statistics

```{r}
head(rowMeans(grading_system[, c("math", "english", "physics")]))
```

--

```{r}
colMeans(grading_system[, c("math", "english", "physics")])
colSums(grading_system[, c("math", "english", "physics")]) / nrow(grading_system)
```

--

**Missing value?**

--

Set `na.rm = TRUE`.

---

## Some simple descriptive statistics

```{r}
var(grading_system[, "math"])
```

--

```{r}
mean((grading_system[, "math"] - mean(grading_system[, "math"]))^2)
```

--

```{r}
sum((grading_system[, "math"] - mean(grading_system[, "math"]))^2) / (nrow(grading_system) - 1)
```

--

```{r}
sd(grading_system[, "math"])
```

--

```{r}
sqrt(var(grading_system[, "math"]))
```

---

## Some simple descriptive statistics

```{r}
knitr::kable(var(grading_system[, c("math", "english", "physics")]), format = "html")
```

--

```{r}
sum((grading_system[, "math"] - mean(grading_system[, "math"])) *
  (grading_system[, "physics"] - mean(grading_system[, "physics"]))) /
  (nrow(grading_system) - 1)
```

---

## Some simple descriptive statistics

```{r}
knitr::kable(cor(grading_system[, c("math", "english", "physics")]), format = "html")
```

**In-class exercise: How to calculate the correlation between `math` and `physics` by yourself?**

---
class: inverse, center, middle

# How about the visualization?

---

## Histogram

```{r fig.align="center",out.width="50%"}
hist(grading_system[, "average"])
```

---

## Boxplot

```{r fig.align="center",out.width="50%"}
boxplot(grading_system[, "average"])
```

---

## Boxplot

```{r fig.align='center',out.width="50%"}
boxplot(grading_system[, c("math", "english", "physics", "average")])
```

---

## Scatter plot

```{r fig.align="center",out.width="50%"}
plot(x = grading_system[, "math"], y = grading_system[, "average"])
```

---

## Scatter plot

```{r fig.align="center",out.width="50%"}
plot(grading_system[, c("gender", "math", "english", "physics", "average")])
```

---

## Scatter plot

```{r fig.align="center",out.width="35%"}
fit_lm <- lm(average ~ math, data = grading_system)
plot(average ~ math, data = grading_system)
abline(fit_lm, col = "red")
fit_lm$coefficients
```

---
class: inverse, center, middle

# Want more beautiful figures?

--

# Use ggplot2!

--

Someone said that if you don’t know ggplot2, it’s like you haven’t learned R!

--

**Actually, "someone" is me:)**

---

## Histogram

.panelset[

```{r panelset=c(source="Code",output="Figure"),fig.align="center",out.width="50%"}
# install.packages("ggplot2")
library(ggplot2)
ggplot(grading_system, aes(x = average)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") +
  geom_vline(aes(xintercept = mean(average)), color = "red", linetype = "dashed", linewidth = 3) +
  labs(title = "Histogram of average score", x = "Average score", y = "Frequency") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 24, hjust = 0.5),
        axis.title = element_text(face = "bold", size = 20),
        axis.text = element_text(size = 12),
        axis.text.x = element_text(size = 16),
        axis.text.y = element_text(size = 16))
```

]

---

## Histogram in group

.panelset[

.panel[.panel-name[`group` argument]

We can also generate the histogram in group.

]

```{r panelset=c(source="Code",output="Figure"),fig.align="center",out.width="50%"}
ggplot(grading_system, aes(x = average, group = group)) +
  geom_histogram(aes(fill = group, color = group)) +
  labs(title = "Histogram of average score", x = "Average score", y = "Frequency") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 24, hjust = 0.5),
        axis.title = element_text(face = "bold", size = 20),
        axis.text = element_text(size = 12),
        axis.text.x = element_text(size = 16),
        axis.text.y = element_text(size = 16),
        legend.title = element_text(size = 16),
        legend.text = element_text(size = 16))
```

]

---

## Histogram with `facet_grid()`

.panelset[

.panel[.panel-name[`facet_grid()`]

`facet_grid()` forms a matrix of panels defined by row and column faceting variables. It is most useful when you have two discrete variables, and all combinations of the variables exist in the data. 

Let's see the following example to generate the histograms of all three subjects in each group.

]

```{r panelset=c(source="Code",output="Figure"),fig.align="center",out.width="50%"}
grading_system_long <- reshape2::melt(grading_system[, -7], 
                                      id = c("id", "group", "gender"), 
                                      variable.name = "subject", 
                                      value.name = "score")
ggplot(grading_system_long, aes(x = score)) +
  geom_histogram(aes(fill = subject, color = subject)) +
  labs(title = "Histogram of all subjects", x = "Score", y = "Frequency") +
  facet_grid(group ~ subject) +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 24, hjust = 0.5),
        axis.title = element_text(face = "bold", size = 20),
        axis.text = element_text(size = 16),
        axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
        axis.text.y = element_text(size = 12),
        legend.title = element_text(size = 16),
        legend.text = element_text(size = 16),
        strip.text = element_text(size = 16, color = "#490573"))
```

]

---

## Boxplot

.panelset[

```{r panelset=c(source="Code",output="Figure"),fig.align="center",out.width="50%"}
ggplot(grading_system, aes(y = average)) +
  geom_boxplot(color = "black", fill = "pink") +
  geom_hline(aes(yintercept = mean(average)), color = "red", linetype = "dashed", linewidth = 2) +
  labs(title = "Boxplot of average score", y = "Frequence") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 24, hjust = 0.5),
        axis.title = element_text(face = "bold", size = 20),
        axis.text = element_text(size = 12),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 16))
```

]

---

## Boxplot in group

.panelset[

```{r panelset=c(source="Code",output="Figure"),fig.align="center",out.width="50%"}
ggplot(grading_system, aes(x = group, y = average)) +
  geom_boxplot(aes(fill = group, color = group)) +
  labs(title = "Boxplot of average score", x = "Group", y = "Average score") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 24, hjust = 0.5),
        axis.title = element_text(face = "bold", size = 20),
        axis.text = element_text(size = 12),
        axis.text.x = element_text(size = 16),
        axis.text.y = element_text(size = 16),
        legend.title = element_text(size = 16),
        legend.text = element_text(size = 16))
```

]

---

## Boxplots with `facet_grid()`

.panelset[

```{r panelset=c(source="Code",output="Figure"),fig.align="center",out.width="50%"}
ggplot(grading_system_long, aes(x = group, y = score)) +
  geom_boxplot(aes(fill = group, color = group)) +
  labs(title = "Boxplots of scores", x = "Group", y = "Frequence") +
  facet_grid(subject ~ .) +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 24, hjust = 0.5),
        axis.title = element_text(face = "bold", size = 20),
        axis.text = element_text(size = 16),
        axis.text.x = element_text(size = 16),
        axis.text.y = element_text(size = 16),
        legend.title = element_text(size = 16),
        legend.text = element_text(size = 16),
        strip.text = element_text(size = 16))
```

]

---

## Violin plot

.panelset[

```{r panelset=c(source="Code",output="Figure"),fig.align="center",out.width="50%"}
ggplot(grading_system_long, aes(x = group, y = score)) +
  geom_violin(aes(fill = group, color = group)) +
  labs(title = "Violin plots of scores", x = "Group", y = "Frequence") +
  facet_grid(subject ~ .) +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 24, hjust = 0.5),
        axis.title = element_text(face = "bold", size = 20),
        axis.text = element_text(size = 16),
        axis.text.x = element_text(size = 16),
        axis.text.y = element_text(size = 16),
        legend.title = element_text(size = 16),
        legend.text = element_text(size = 16),
        strip.text = element_text(size = 16))
```

]

---

## Violin plot with boxplot inner

.panelset[

```{r panelset=c(source="Code",output="Figure"),fig.align="center",out.width="50%",warning=FALSE}
ggplot(grading_system_long, aes(x = group, y = score)) +
  geom_violin(aes(fill = group), trim = FALSE) +
  geom_boxplot(width = 0.1, fill = "white", color = "black") +
  ylim(50, 100) +
  labs(title = "Violin plots of scores", x = "Group", y = "Frequence") +
  facet_grid(subject ~ .) +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 24, hjust = 0.5),
        axis.title = element_text(face = "bold", size = 20),
        axis.text = element_text(size = 16),
        axis.text.x = element_text(size = 16),
        axis.text.y = element_text(size = 16),
        legend.title = element_text(size = 16),
        legend.text = element_text(size = 16),
        strip.text = element_text(size = 16))
```

]   

---

## Scatter plot

.panelset[

```{r panelset=c(source="Code",output="Figure"),fig.align="center",out.width="50%"}
# Use only first two groups
ggplot(grading_system[grading_system$group %in% c(1, 2), ], aes(x = math, y = physics)) +
  facet_grid(. ~ group) +
  geom_point(aes(color = group), size = 3) +
  geom_smooth(method = "lm") +
  labs(title = "Scatter plot of math and physics", x = "Math", y = "Physics") +
  coord_fixed(ratio = 1) +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 24, hjust = 0.5),
        axis.title = element_text(face = "bold", size = 20),
        axis.text = element_text(size = 16),
        axis.text.x = element_text(size = 16),
        axis.text.y = element_text(size = 16),
        legend.title = element_text(size = 16),
        legend.text = element_text(size = 16),
        strip.text = element_text(size = 16))
```

]

---

## Simulate a more reasonable dataset 

.panelset[

.panel[.panel-name[Multivariate normal distribution]

```{r}
mean_vec <- c(70, 70)
cov_mat <- matrix(c(25, 15, 15, 25), nrow = 2)
math_physics <- MASS::mvrnorm(n = n_students, 
                              mu = mean_vec, Sigma = cov_mat)
math_physics <- round(math_physics)
math_physics[math_physics < 55] <- 55
math_physics[math_physics > 95] <- 95
grading_system[, c("math", "physics")] <- math_physics
```

]

```{r panelset=c(source="Code",output="Figure"),fig.align="center",out.width="50%",warning=FALSE}
ggplot(grading_system[grading_system$group %in% c(1, 2), ], aes(x = math, y = physics)) +
  facet_grid(. ~ group) +
  geom_point(aes(color = group), size = 3) +
  geom_smooth(method = "lm") +
  labs(title = "Scatter plot of math and physics", x = "Math", y = "Physics") +
  coord_fixed(ratio = 1) +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 24, hjust = 0.5),
        axis.title = element_text(face = "bold", size = 20),
        axis.text = element_text(size = 16),
        axis.text.x = element_text(size = 16),
        axis.text.y = element_text(size = 16),
        legend.title = element_text(size = 16),
        legend.text = element_text(size = 16),
        strip.text = element_text(size = 16))
```

]

---
class: inverse, center, middle

# Thank you!

Slides created via Yihui Xie's R package [**xaringan**](https://github.com/yihui/xaringan).

Theme customized via Garrick Aden-Buie's R package [**xaringanthemer**](https://github.com/gadenbuie/xaringanthemer).

Tabbed panels created via Garrick Aden-Buie's R package [**xaringanExtra**](https://github.com/gadenbuie/xaringanExtra/).

The chakra comes from [remark.js](https://remarkjs.com), [**knitr**](https://yihui.org/knitr/), and [R Markdown](https://rmarkdown.rstudio.com).