---
title: "T04: Random Variables"
subtitle: "MATH	2411 Applied Statistics"
author: "WANG Zhiwei"
institute: "MATH, HKUST"
date: "2024-09-30"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  results = "asis",
  echo = TRUE,
  comment = "#>",
  out.width = "100%"
)
library(xaringanthemer)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
style_mono_accent(
  header_font_google = google_font("Josefin Slab", "600"),
  text_font_google   = google_font("Work Sans", "300", "300i"),
  code_font_google   = google_font("IBM Plex Mono")
)

style_mono_light(
  base_color = "#003366",
  link_color = "#996600",
  text_bold_color = "#996600",
  link_decoration = "underline"
)
```

```{r xaringan-panelset, echo=FALSE}
xaringanExtra::use_panelset()
```

class: inverse, center, middle

# There are two common types of r.v.'s.

We first talk about the discrete random variable.

---

## Discrete Random Variable

Let $X$ be a discrete random variable:

- $p(x)$, the Probability Mass Function (PMF) of $X$, is the probability that event $X = x$ will occur for each $x$ in the range of $X$, i.e., $p(x) = P(X = x)$.

- $F(x)$, the Cumulative Distribution Function (CDF) of $X$, is defined as $F(x) = P(X \leq x)$.

- $E(X)$, the Expectation of $X$, is defined as
  $$E(X) = \sum_{x \in \text{Range}(X)} [x \cdot P(x)]$$

- $\text{Var}(X)$, the Variance of $X$, is defined as
  $$\text{Var}(X) = \sum_{x \in \text{Range}(X)} [(x - E(X))^2 \cdot P(x)] = E((X - E(X))^2)$$

---

## Proporties of population mean and variance

When it exists, the mathematical expectation $E$ satisfies the following properties: Suppose $X, Y$ are random variables and $a$ and $b$ are two constants. Then

- $E(b) = b$
- $E(aX) = aE(X)$
- $E(aX + b) = aE(X) + b$
- $E(X + Y) = E(X) + E(Y)$

--

When it exists, the population variance satisfies the following properties: Suppose $X$ and $Y$ are random variables and $a$ and $b$ are two constants. Then

- $\text{Var}(b) = 0$
- $\text{Var}(aX) = a^2 \text{Var}(X)$
- $\text{Var}(aX + b) = a^2 \text{Var}(X)$
- If $X$ and $Y$ are independent random variables, then $\text{Var}(X + Y) = \text{Var}(X) + \text{Var}(Y)$.


---

### Problem 1.1

Let $W$ be a random variable giving the number of heads minus the number of tails in four tosses of a coin. List the elements of the sample space $S$ for the four tosses of the coin and to each sample point assign a value $w$ of $W$.

--

  **Solution** Let 'H' denote head and 'T' denote tail. The sample space $S$ and the random variable $W$ are as follows:

| $S$                               | $W$ |
|-----------------------------------|-----|
| HHHH                              | 4   |
| HHHT, HHTH, HTHH, THHH            | 2   |
| HHTT, HTHT, HTTH, THHT, THTH, TTHH| 0   |
| HTTT, THTT, TTHT, TTTH            | -2  |
| TTTT                              | -4  |

---

### Problem 1.2

A coin is flipped until 2 heads occur in succession. List only those elements of the sample space that require 6 or fewer tosses. Is this a discrete sample space? Explain.

--

  **Solution** Let 'H' denote head and 'T' denote tail. The possible elements are:

- HH
- THH
- TTHH, HTHH
- HTTHH, TTTTH, THTHH
- TTTTHH, HTTTHH, THTTHH, TTHTHH, HTHTHH

The sample space is discrete since it contains finite elements.


---

### Problem 1.3

Let $X$ be a random variable with the following probability distribution:

| $x$ | 2   | 3   | 4   |
|-----|-----|-----|-----|
| $p(x)$ | $\frac{4}{15}$ | $\frac{4}{15}$ | $\frac{7}{15}$ |

--

a. Find Expected Value $E(X)$

--

b. Find Variance $\text{Var}(X)$

--

c. Find Standard Deviation $\sigma(Y)$, where $Y = 2X - 1$ (Hint: $\sigma(X) = \sqrt{\text{Var}(X)}$)
   
--

d. Find the CDF of $X$ and plot the graph of the CDF.

---

### Problem 1.4

Given that the cdf (cumulative distribution function) of a discrete random variable $X$ is 

$$\begin{align*}
F(x) = 
\begin{cases}
0, & x < 1 \\
\frac{1}{4}, & 1 \leq x < 3 \\
\frac{3}{4}, & 3 \leq x < 5 \\
1, & x \geq 5
\end{cases}
\end{align*}$$

i) Find the pmf (probability mass function) of $X$.

ii) Draw the graphs of cdf and pmf of $X$.

iii) Evaluate $P(X > 3)$ and $P(1.5 < X < 5)$.

iv) Let $Y = 3X - 1$, find the population mean and population variance of $Y$.

---
class: inverse, center, middle

# Some probability distributions of discrete random variables.

---

## Bernoulli distribution

A Bernoulli random variable takes on only two values: $0$ and $1$, with probabilities $1 - p$ and $p$, respectively. Its PMF is thus

$$\begin{align*}
& p(1) = p \\
& p(0) = 1 - p \\
& p(x) = 0, \quad \text{if } x \neq 0 \text{ and } x \neq 1.
\end{align*}$$

--

An alternative and sometimes useful representation of this function is 
$$\begin{align*}
p(x) = 
\begin{cases} 
p^x(1 - p)^{1-x}, & \text{if } x = 0 \text{ or } x = 1 \\
0, & \text{otherwise}
\end{cases}
\end{align*}$$

--

**Typical examples:** (If an experiment/trial has only **two possible outcomes**, it can be described by a Bernoulli r.v.)
- Tossing a coin: “head” or “tail”
- The grade of a P/F course: “Pass” or “Fail”

--

**Questions:** CDF? $E(X)$ and $Var(X)$?

---

## Binomial distribution

- If $X$ is the discrete r.v. of the **number of successes in $n$ trials**, then we can use a **Binomial distribution** to characterize its random behavior.
 
--
 
- Here are some typical examples:
  - The number of heads in $n$ tosses of a coin.
  - The number of defective products in a sample of $n$ products.
  - The number of students who pass a test in a class of $n$ students.
  - ......

--

- A Binomial distribution is related to a random experiment with the following features:
 - Fixed **finite** number of identical trials, say $n < \infty$.
 - Trials are **independent**.
 - Trials result in **two possible outcomes** denoted by success and failure.
 
---

## Binomial distribution

- **Notation** 
  $$ X \sim \text{Binomial}(n, p),$$
  where $n$: number of trials; $p$: probability of success.
  
--

- **PMF**
  $$\begin{align*}
  & p(x) = P(X = x) = \binom{n}{x} p^x (1 - p)^{n-x}, \quad \text{for } x = 0, 1, \ldots, n \\
  & p(x) = 0, \quad \text{otherwise.}
\end{align*}$$

**Remark:** The total number of ways that $x$ successes occur
in $n$ trials is $\binom{n}{x}$.

--

- **Mean and Variance**
  $$
  E(X) = np, \quad  Var(X) = np(1-p)
  $$

---

## Poisson distribution

The **Poisson distribution** can be used to model the **number of events occurring in a fixed interval of time or space**.

Here are some typical examples for Poisson distributions:
- The number of traffic accidents occurring on a highway in a day.
- Crashes of a computer network per week.
- The number of people joining a line in an hour.
- The number of customers arrived per day.
- The number of goals scored in a hockey game.
- The number of typos per page of an essay.
- The number of mutations on a DNA strand.
- ......

---

## Poisson distribution

- **Notation** Let $X$ be the number of occurrences of an event over a unit time. Then we say that $X$ follows a Poisson distribution, denoted by
  $$X \sim Poisson(\lambda),$$

  where $\lambda \in (0, \infty)$ is the **rate of occurrences** of an event **per unit time (or space)** or the **average number of occurrences** of the event **per unit time (or space)**.

- **PMF**
  $$\begin{align*}
  & p(x) = P(X = x) = \frac{\lambda^x e^{-\lambda}}{x!}, \quad \text{for } x = 0, 1, \ldots \\
  & p(x) = 0, \quad \text{otherwise.}
  \end{align*}$$

**Remark**
$\sum_{x=0}^{\infty} P\{X = x\} = \sum_{x=0}^{\infty} \frac{\lambda^x e^{-\lambda}}{x!} = e^{-\lambda} \cdot \sum_{x=0}^{\infty} \frac{\lambda^x}{x!} = e^{-\lambda} \cdot e^{\lambda} = 1$ (Using Maclaurin series of $e^{\lambda}$).

---

### Problem 2.1

The probability that a patient recovers from a delicate heart operation is $0.8$. What is the probability that

i) exactly 2 of the next 3 patients who have this operation survive?

--

  **Solution**

  Let $P(A_i) = 0.8$ for $i = 1, 2, 3$. The probability that a patient does not survive is $P(A_i^c) = 0.2$.

The probability that exactly 2 out of 3 patients survive is:

$$\binom{3}{2} \cdot (0.8)^2 \cdot (0.2) = 3 \cdot 0.64 \cdot 0.2 = 0.384$$

--

ii) all of the next 3 patients who have this operation survive?

--

  **Solution**
  
  The probability that all 3 patients survive is:

$$(0.8)^3 = 0.512$$

---

### Problem 2.2

There are 8 female students and 24 male students in MATH2411 Tutorial. For the next 5 weeks, every week after class one student will be chosen randomly to invite everybody for tea. Let $X$ be the total number of female students chosen.

--

i) Find the probability that the 'paying gender' sequence will be FMMFM.

--

  **Solution**
  $$\frac{1}{4} \cdot \frac{3}{4} \cdot \frac{3}{4} \cdot \frac{1}{4} \cdot \frac{3}{4}$$

--

ii) Find the probability that the 'paying gender' sequence will be MMFFM.

--

  **Solution**
  $$\frac{3}{4} \cdot \frac{3}{4} \cdot \frac{1}{4} \cdot \frac{1}{4} \cdot \frac{3}{4}$$

---

### Problem 2.2

iii) In how many ways 2 out of the 5 places can be chosen?

--

  **Solution**
  $$\binom{5}{2} = 10$$

--

iv) What is the probability that 2 out of the 5 weeks will be paid by female students?

--

  **Solution**
  $$\binom{5}{2} \cdot \left(\frac{1}{4}\right)^2 \cdot \left(\frac{3}{4}\right)^3$$

---

### Problem 2.2

v) What is the probability that no more than 1 week will be paid by female students?

--

  **Solution** 
  $$P(X \leq 1) = P(X = 0) + P(X = 1) = \left(\frac{3}{4}\right)^5 + \binom{5}{1} \cdot \left(\frac{1}{4}\right) \cdot \left(\frac{3}{4}\right)^4$$

--

vi) What is the probability that at least 1 week will be paid by a male student?

--

  **Solution**
  $$P(X < 5) = 1 - P(X = 5) = 1 - \left(\frac{1}{4}\right)^5$$

--

vii) What is the mean of $X$?

--

  **Solution**
  $$E(X) = 5 \cdot \frac{1}{4}$$
  
---

### Problem 2.3

In California, we are monitoring earthquakes that occur at a magnitude greater than 6.7 on the Richter scale. Using the Poisson distribution with the average rate $\lambda = 1.5$ to model the number of such earthquakes per year, we would like to find the following probability:

i) What is the probability of having 3 or more occurrences in the upcoming year?

--

  **Solution**
  $$P(X \geq 3) = 1 - P(X < 3) = 1 - \left(\frac{e^{-1.5} \cdot 1.5^0}{0!} + \frac{e^{-1.5} \cdot 1.5^1}{1!} + \frac{e^{-1.5} \cdot 1.5^2}{2!}\right)$$
  
---

### Problem 2.3

ii) What is the probability of having 3 or more occurrences in the next 5 years?

--

  **Solution**
  Since it covers 5 years, it is equivalent to $P(X \geq 3)$ when $X \sim \text{Poisson}(7.5)$, where the rate parameter comes from $1.5 \times 5 = 7.5$. Therefore,
  $$P(X \geq 3) = 1 - P(X < 3) = 1 - \left(\frac{e^{-7.5} \cdot 7.5^0}{0!} + \frac{e^{-7.5} \cdot 7.5^1}{1!} + \frac{e^{-7.5} \cdot 7.5^2}{2!}\right)$$
  
--

iii) Given that there are 52 weeks in a year, what is the probability of going 1 week without an earthquake over 6.7?

--

  **Solution**
  In one week, the number of such earthquakes follows $X \sim \text{Poisson}(0.0288)$, where the rate parameter comes from $\frac{1.5}{52} = 0.0288$. Then we have:

$$
P(X = 0) = e^{-0.0288} = 0.9716.
$$

---

### Problem 2.3

iv) What is the probability of going 18 months without an earthquake over 6.7?

--

  **Solution**
  In 18 months, the number of such earthquakes follows $X \sim \text{Poisson}(2.25)$, where the rate parameter comes from $1.5 \times \frac{18}{12} = 2.25$. Then we have:

$$
P(X = 0) = e^{-2.25} = 0.1054.
$$

--

v) What is the expected number of earthquakes over 6.7 in the next 4 years?

--

  **Solution**
  In the next 4 years, the number of such earthquakes follows $X \sim \text{Poisson}(6)$. Therefore, the expected number is 6.
  
---

### Problem 2.4

Let the random variable $X \sim \text{Poisson}(\lambda)$. Show that 

$$
E(X) = \lambda \quad \text{and} \quad Var(X) = \lambda.
$$

--

  **Solution** Recall that the pdf of $X$ is 

$$
f(x) = \frac{e^{-\lambda} \lambda^x}{x!} \quad \text{for integers } x = 0, 1, 2, \ldots
$$ 

Therefore, by definition, we have

$$\begin{align*}
E(X) = & \sum_{x=0}^{\infty} x f(x) = e^{-\lambda} \sum_{x=1}^{\infty} \frac{x \lambda^x}{x!} \\
= & e^{-\lambda} \sum_{x=1}^{\infty} \frac{\lambda^{x}}{(x-1)!} = \lambda e^{-\lambda} \sum_{x=0}^{\infty} \frac{\lambda^{x}}{x!} \\
= & \lambda e^{-\lambda} \cdot e^{\lambda} = \lambda.
\end{align*}$$

---

### Problem 2.4

Similarly, we have

$$\begin{align*}
E(X^2) = & e^{-\lambda} \sum_{x=1}^{\infty} \frac{x^2 \lambda^x}{x!} = e^{-\lambda} \sum_{x=1}^{\infty} \frac{x (x-1) \lambda^x}{(x-1)!} + e^{-\lambda} \sum_{x=1}^{\infty} \frac{x \lambda^x}{(x-1)!} \\
= & e^{-\lambda} \sum_{x=2}^{\infty} \frac{\lambda^{x}}{(x-2)!} + \lambda e^{-\lambda} \sum_{x=0}^{\infty} \frac{\lambda^{x}}{x!} \\
= & \lambda^2 e^{-\lambda} \sum_{x=0}^{\infty} \frac{\lambda^{x}}{x!} + \lambda e^{-\lambda} \sum_{x=0}^{\infty} \frac{\lambda^{x}}{x!} = \lambda^2 + \lambda.
\end{align*}$$

As a result, we get

$$
Var(X) = E(X^2) - (E(X))^2 = \lambda^2 + \lambda - \lambda^2 = \lambda.
$$

---

### Problem 2.5

[Please answer the question and then use R to check your answer, and provide your R codes.]

The probability that a patient recovers from a rare blood disease is $p = 0.004$. If 1500 people are known to have contracted this disease.

i) What is the probability that
  (a) at least 10 survive?
  (b) from 3 to 8 survive?
  (c) exactly 5 survive?
  
--

  **Solution** The binomial pmf is given by:
$$
P(X = k) = \binom{n}{k} p^k (1-p)^{n-k},
$$
where $n = 1500$ and $p = 0.004$.

(a) $P(X \geq 10) = 1 - P(X \leq 9) = 1 - \sum_{k=0}^{9} P(X = k)$.

(b) $P(3 \leq X \leq 8) = P(X \leq 8) - P(X \leq 2)$.

(c) $P(X = 5)$.

---

### Problem 2.5

**R code**

.panelset[

.panel[.panel-name[Set up parameters]

```{r}
n <- 1500
p <- 0.004
```

]

.panel[.panel-name[(a)]

```{r}
## Use pmf
prob_a <- 1
for (k in 0:9) {
  prob_a <- prob_a - dbinom(x = k, size = n, prob = p)
}
prob_a
```

```{r}
## Use cdf
prob_a <- 1 - pbinom(q = 9, size = n, prob = p)
prob_a
```

]

.panel[.panel-name[(b)]

```{r}
## Use pmf
prob_b <- 0
for (k in 3:8) {
  prob_b <- prob_b + dbinom(x = k, size = n, prob = p)
}
prob_b
```

```{r}
## Use cdf
prob_b <- pbinom(q = 8, size = n, prob = p) - 
  pbinom(q = 2, size = n, prob = p)
prob_b
```

]

.panel[.panel-name[(c)]

```{r}
## Use pmf
prob_c <- dbinom(x = 5, size = n, prob = p)
prob_c
```

```{r}
## Use cdf
prob_c <- pbinom(q = 5, size = n, prob = p) - 
  pbinom(q = 4, size = n, prob = p)
prob_c
```

]

]

---

## The relationship between Binomial & Poisson distributions

The binomial distribution tends towards the Poisson distribution as $n \to \infty$, $p \to 0$ and $np$ stays constant.

### Poisson Limit Theorem

(Poisson approximation to Binomial Distribution)

Let $p_n$ be a sequence of real numbers in $[0,1]$ such that the sequence $n p_n$ converges to a finite limit $\lambda$. Then:
$$
\lim_{n \to \infty} \binom{n}{k} p_n^k (1 - p_n)^{n-k} = \frac{\lambda^k e^{-\lambda}}{k!}
$$
**The Poisson distribution with $\lambda = np$ closely approximates the binomial distribution if $n$ is large and $p$ is small.**

---

### Problem 2.5

ii) Use Poisson distribution to approximate the probabilities above.

--

  **Solution** The Poisson probability mass function is given by:
$$
P(X = k) = \frac{\lambda^k e^{- \lambda}}{k!},
$$
where the rate parameter is $\lambda = np = 1500 \times 0.004 = 6$.

Then we can replace the binomial pmf in i) with the Poisson pmf to approximate the probabilities.

---

### Problem 2.5

**R code**

.panelset[

.panel[.panel-name[Set up parameters]

```{r}
n <- 1500
p <- 0.004
lambda <- n * p
```

]

.panel[.panel-name[(a)]

```{r}
## Use pmf
prob_a <- 1
for (k in 0:9) {
  prob_a <- prob_a - dpois(x = k, lambda = lambda)
}
prob_a
```

```{r}
## Use cdf
prob_a <- 1 - ppois(q = 9, lambda = lambda)
prob_a
```

]

.panel[.panel-name[(b)]

```{r}
## Use pmf
prob_b <- 0
for (k in 3:8) {
  prob_b <- prob_b + dpois(x = k, lambda = lambda)
}
prob_b
```

```{r}
## Use cdf
prob_b <- ppois(q = 8, lambda = lambda) - 
  ppois(q = 2, lambda = lambda)
prob_b
```

]

.panel[.panel-name[(c)]

```{r}
## Use pmf
prob_c <- dpois(x = 5, lambda = lambda)
prob_c
```

```{r}
## Use cdf
prob_c <- ppois(q = 5, lambda = lambda) - 
  ppois(q = 4, lambda = lambda)
prob_c
```

]

]

---

### Problem 2.6

Assume that the number of flu a person gets in a year follows a Poisson distribution with parameter $\lambda = 5$.
Now a medicine to prevent flu is developed and tested.
Suppose that the medicine is effective for 75% of the people, by reducing the parameter of the Poisson distribution to 3; but it does not work for the other 25% of the people.
If someone takes this medicine and gets flu twice in one year, what is the probability that the medicine is effective for this person?

---
class: inverse, center, middle

# Thank you!

Slides created via Yihui Xie's R package [**xaringan**](https://github.com/yihui/xaringan).

Theme customized via Garrick Aden-Buie's R package [**xaringanthemer**](https://github.com/gadenbuie/xaringanthemer).

Tabbed panels created via Garrick Aden-Buie's R package [**xaringanExtra**](https://github.com/gadenbuie/xaringanExtra/).

The chakra comes from [remark.js](https://remarkjs.com), [**knitr**](https://yihui.org/knitr/), and [R Markdown](https://rmarkdown.rstudio.com).