---
title: "T04: Random Variables"
subtitle: "MATH	2411 Applied Statistics"
author: "WANG Zhiwei"
institute: "MATH, HKUST"
date: "2024-09-30"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  results = "asis",
  echo = TRUE,
  comment = "#>",
  out.width = "100%"
)
library(xaringanthemer)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
style_mono_accent(
  header_font_google = google_font("Josefin Slab", "600"),
  text_font_google   = google_font("Work Sans", "300", "300i"),
  code_font_google   = google_font("IBM Plex Mono")
)

style_mono_light(
  base_color = "#003366",
  link_color = "#996600",
  text_bold_color = "#996600",
  link_decoration = "underline"
)
```

```{r xaringan-panelset, echo=FALSE}
xaringanExtra::use_panelset()
```

class: inverse, center, middle

# Two common types of r.v.'s

---

## Discrete Random Variable

Let $X$ be a discrete random variable:

- $p(x)$, the Probability Mass Function (PMF) of $X$, is the probability that event $X = x$ will occur for each $x$ in the range of $X$, i.e., $p(x) = P(X = x)$.

- $F(x)$, the Cumulative Distribution Function (CDF) of $X$, is defined as $F(x) = P(X \leq x)$.

- $E(X)$, the Expectation of $X$, is defined as
  $$E(X) = \sum_{x \in \text{Range}(X)} [x \cdot P(x)]$$

- $\text{Var}(X)$, the Variance of $X$, is defined as
  $$\text{Var}(X) = \sum_{x \in \text{Range}(X)} [(x - E(X))^2 \cdot P(x)] = E((X - E(X))^2)$$

---

## Proporties of population mean and variance

When it exists, the mathematical expectation $E$ satisfies the following properties: Suppose $X, Y$ are random variables and $a$ and $b$ are two constants. Then

- $E(b) = b$
- $E(aX) = aE(X)$
- $E(aX + b) = aE(X) + b$
- $E(X + Y) = E(X) + E(Y)$

--

When it exists, the population variance satisfies the following properties: Suppose $X$ and $Y$ are random variables and $a$ and $b$ are two constants. Then

- $\text{Var}(b) = 0$
- $\text{Var}(aX) = a^2 \text{Var}(X)$
- $\text{Var}(aX + b) = a^2 \text{Var}(X)$
- If $X$ and $Y$ are independent random variables, then $\text{Var}(X + Y) = \text{Var}(X) + \text{Var}(Y)$.

---

## Problem 1

Let $X$ be a random variable with the following probability distribution:

| $x$ | 2   | 3   | 4   |
|-----|-----|-----|-----|
| $p(x)$ | $\frac{4}{15}$ | $\frac{4}{15}$ | $\frac{7}{15}$ |

--

a. Find Expected Value $E(X)$

--

b. Find Variance $\text{Var}(X)$

--

c. Find Standard Deviation $\sigma(Y)$, where $Y = 2X - 1$ (Hint: $\sigma(X) = \sqrt{\text{Var}(X)}$)
   
--

d. Find the CDF of $X$ and plot the graph of the CDF.

---

## Problem 2

There are 8 female students and 24 male students in MATH2411 Tutorial. For the next 5 weeks, every week after class one student will be chosen randomly to invite everybody for tea. Let $X$ be the total number of female students chosen.

--

a. Find the probability that the 'paying gender' sequence will be FMMFM.

--

b. Find the probability that the 'paying gender' sequence will be MMF FM.

--

c. In how many ways 2 out of the 5 places can be chosen?

--

d. What is the probability that 2 out of the 5 weeks will be paid by female students?

--

e. What is the probability that no more than 1 week will be paid by female students?

--

f. What is the probability that at least 1 week will be paid by a male student?

--

g. What is the mean of $X$?

---
class: inverse, center, middle

# Thank you!

Slides created via Yihui Xie's R package [**xaringan**](https://github.com/yihui/xaringan).

Theme customized via Garrick Aden-Buie's R package [**xaringanthemer**](https://github.com/gadenbuie/xaringanthemer).

Tabbed panels created via Garrick Aden-Buie's R package [**xaringanExtra**](https://github.com/gadenbuie/xaringanExtra/).

The chakra comes from [remark.js](https://remarkjs.com), [**knitr**](https://yihui.org/knitr/), and [R Markdown](https://rmarkdown.rstudio.com).