---
title: "T04: Random Variables"
subtitle: "MATH	2411 Applied Statistics"
author: "WANG Zhiwei"
institute: "MATH, HKUST"
date: "2024-09-30"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  results = "asis",
  echo = TRUE,
  comment = "#>",
  out.width = "100%"
)
library(xaringanthemer)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
style_mono_accent(
  header_font_google = google_font("Josefin Slab", "600"),
  text_font_google   = google_font("Work Sans", "300", "300i"),
  code_font_google   = google_font("IBM Plex Mono")
)

style_mono_light(
  base_color = "#003366",
  link_color = "#996600",
  text_bold_color = "#996600",
  link_decoration = "underline"
)
```

```{r xaringan-panelset, echo=FALSE}
xaringanExtra::use_panelset()
```

class: inverse, center, middle

# Two common types of r.v.'s

---

## Discrete Random Variable

Let $X$ be a discrete random variable:

- $p(x)$, the Probability Mass Function (PMF) of $X$, is the probability that event $X = x$ will occur for each $x$ in the range of $X$, i.e., $p(x) = P(X = x)$.

- $F(x)$, the Cumulative Distribution Function (CDF) of $X$, is defined as $F(x) = P(X \leq x)$.

- $E(X)$, the Expectation of $X$, is defined as
  $$E(X) = \sum_{x \in \text{Range}(X)} [x \cdot P(x)]$$

- $\text{Var}(X)$, the Variance of $X$, is defined as
  $$\text{Var}(X) = \sum_{x \in \text{Range}(X)} [(x - E(X))^2 \cdot P(x)] = E((X - E(X))^2)$$

---

## Proporties of population mean and variance

When it exists, the mathematical expectation $E$ satisfies the following properties: Suppose $X, Y$ are random variables and $a$ and $b$ are two constants. Then

- $E(b) = b$
- $E(aX) = aE(X)$
- $E(aX + b) = aE(X) + b$
- $E(X + Y) = E(X) + E(Y)$

--

When it exists, the population variance satisfies the following properties: Suppose $X$ and $Y$ are random variables and $a$ and $b$ are two constants. Then

- $\text{Var}(b) = 0$
- $\text{Var}(aX) = a^2 \text{Var}(X)$
- $\text{Var}(aX + b) = a^2 \text{Var}(X)$
- If $X$ and $Y$ are independent random variables, then $\text{Var}(X + Y) = \text{Var}(X) + \text{Var}(Y)$.


---

### Problem 1.1

Let $W$ be a random variable giving the number of heads minus the number of tails in four tosses of a coin. List the elements of the sample space $S$ for the four tosses of the coin and to each sample point assign a value $w$ of $W$.

--

  **Solution** Let 'H' denote head and 'T' denote tail. The sample space $S$ and the random variable $W$ are as follows:

| $S$                               | $W$ |
|-----------------------------------|-----|
| HHHH                              | 4   |
| HHHT, HHTH, HTHH, THHH            | 2   |
| HHTT, HTHT, HTTH, THHT, THTH, TTHH| 0   |
| HTTT, THTT, TTHT, TTTH            | -2  |
| TTTT                              | -4  |

---

### Problem 1.2

A coin is flipped until 2 heads occur in succession. List only those elements of the sample space that require 6 or fewer tosses. Is this a discrete sample space? Explain.

--

  **Solution** Let 'H' denote head and 'T' denote tail. The possible elements are:

- HH
- THH
- TTHH, HTHH
- HTTHH, TTTTH, THTHH
- TTTTHH, HTTTHH, THTTHH, TTHTHH, HTHTHH

The sample space is discrete since it contains finite elements.


---

### Problem 1.3

Let $X$ be a random variable with the following probability distribution:

| $x$ | 2   | 3   | 4   |
|-----|-----|-----|-----|
| $p(x)$ | $\frac{4}{15}$ | $\frac{4}{15}$ | $\frac{7}{15}$ |

--

a. Find Expected Value $E(X)$

--

b. Find Variance $\text{Var}(X)$

--

c. Find Standard Deviation $\sigma(Y)$, where $Y = 2X - 1$ (Hint: $\sigma(X) = \sqrt{\text{Var}(X)}$)
   
--

d. Find the CDF of $X$ and plot the graph of the CDF.

---

### Problem 1.4

Given that the cdf (cumulative distribution function) of a discrete random variable $X$ is 

$$
F(x) = 
\begin{cases}
0, & x < 1 \\
\frac{1}{4}, & 1 \leq x < 3 \\
\frac{3}{4}, & 3 \leq x < 5 \\
1, & x \geq 5
\end{cases}
$$

i) Find the pmf (probability mass function) of $X$.

ii) Draw the graphs of cdf and pmf of $X$.

iii) Evaluate $P(X > 3)$ and $P(1.5 < X < 5)$.

iv) Let $Y = 3X - 1$, find the population mean and population variance of $Y$.

---

### Problem 2.1

The probability that a patient recovers from a delicate heart operation is $0.8$. What is the probability that

(a) exactly 2 of the next 3 patients who have this operation survive?

--

  **Solution**

  Let $P(A_i) = 0.8$ for $i = 1, 2, 3$. The probability that a patient does not survive is $P(A_i^c) = 0.2$.

The probability that exactly 2 out of 3 patients survive is:

$$\binom{3}{2} \cdot (0.8)^2 \cdot (0.2) = 3 \cdot 0.64 \cdot 0.2 = 0.384$$

--

(b) all of the next 3 patients who have this operation survive?

  **Solution**
  
  The probability that all 3 patients survive is:

$$(0.8)^3 = 0.512$$

---

## Problem 2.2

There are 8 female students and 24 male students in MATH2411 Tutorial. For the next 5 weeks, every week after class one student will be chosen randomly to invite everybody for tea. Let $X$ be the total number of female students chosen.

--

a. Find the probability that the 'paying gender' sequence will be FMMFM.

--

b. Find the probability that the 'paying gender' sequence will be MMFFM.

--

c. In how many ways 2 out of the 5 places can be chosen?

--

d. What is the probability that 2 out of the 5 weeks will be paid by female students?

--

e. What is the probability that no more than 1 week will be paid by female students?

--

f. What is the probability that at least 1 week will be paid by a male student?

--

g. What is the mean of $X$?

---
class: inverse, center, middle

# Thank you!

Slides created via Yihui Xie's R package [**xaringan**](https://github.com/yihui/xaringan).

Theme customized via Garrick Aden-Buie's R package [**xaringanthemer**](https://github.com/gadenbuie/xaringanthemer).

Tabbed panels created via Garrick Aden-Buie's R package [**xaringanExtra**](https://github.com/gadenbuie/xaringanExtra/).

The chakra comes from [remark.js](https://remarkjs.com), [**knitr**](https://yihui.org/knitr/), and [R Markdown](https://rmarkdown.rstudio.com).